Enriching Object Detection with 2D-3D Registration and Continuous Viewpoint

### 2014-11-14 egpaper_final.tex
fixed bad math
second pass over covariance generation

@@

Recently, Hariharan \etal. introduced Whitened Histograms of Orientations (WHO)
\cite{Hariharan12}, which uses statistics from natural images to whiten HOG
templates, making them more discriminative. Whitening is a common signal
processing operation for decorrelating a set of random variables
\cite{Martinsson05, Belouchrani00}. More formally, suppose that we have a
$k$-dimensional random variable $X \in \mathcal{R}^k$ with $Cov(X)=\Sigma$. By
whitening the signal, i.e., $\tilde{X}=\Sigma^{-\frac{1}{2}}(X - E[X])$, we
remove 2nd order correlation between all features. So, all elements in the
random variable $\tilde{X} = \left[\tilde{X}_1, \tilde{X}_2, \dots,
\tilde{X}_k\right]$, become more independent. In the context of HOG features, a
whitened HOG template will produce Gaussian noise when convolved with a random
natural image, but a delta-like response on an image matching the template. WHO
templates are more discriminative than standard HOG templates because they only
capture the gradients that differentiate the object from random images.

@@

We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), that are both more discriminative than WHO, and can be generated several orders of magnitude faster (in 70ms compared to several seconds for WHO). This speed up means we can generate templates on the fly, allowing our approach to evaluate arbitrary viewpoint hypotheses. 
%\scream{chris:We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), an improvement over WHO which is both more discriminative than WHO and can be computed many orders of magnitude faster than WHO (70ms as opposed to to several seconds for WHO). This speedup allows us to generate NZ-WHO templates on the fly, allowing us to quickly evaluate arbitrary viewpoint hypotheses.}

@@

Our second innovation is 'Non-Zero' whitening. When synthesizing detection templates from rendered images, a common problem is how to handle the background. If we render the model over a natural image background, gradients in the background will be incorporated into the discriminative template. Alternatively, we could leave the background textureless \scream{did Hariharan do this?} (see Fig.~\ref{fig:rendering}), but whitening the resulting HOG template produces in undesirable background artifacts. These are created when centering the template (by subtracting the mean $\mu$), where strong negative weights are introduced in the textureless region (as seen in Fig.~\ref{fig:whocomparison}). This could result in positive matches being suppressed due to spurious background gradients.

To remove such artifacts, we construct $x$ using only cells with non-zero gradients. After whitening we restore the full template, with the zero-gradient cells left unchanged. This approach has the added benefit of reducing the size of the linear system we need to solve, further speeding up our method.

Table \ref{tab:who_initializations} provides a detailed comparison of WHO and NZ-WHO templates, illustrating that our method is faster and more discriminative than Whitened Histograms of Orientations. \scream{Does the table show this?}
% We compare the performance of WHO and NZ-WHO on 3D Object dataset \cite{Savarese07} and report performance on Table \ref{tab:who_initializations}.

@@

% \subsection{Whitening and Textureless Background}

\subsection{Constructing the Covariance Matrix}
\label{sec:feature_statistics}
% Whats used in our work and using Autocovariance matrix
We first collected the first and second order statistics (mean and
autocovariance) of HOG features from arbitrary natural images. Following
\cite{Hariharan12}, we assume Wide-Sense Stationarity (WSS) of HOG features
generated from natural images. That is, we assume the mean of a HOG cell is
independent of its place in the image, and the autocovariance of patches depends
only on their relative location.

>>>>

\end{comment}
\subsection{Rendering Pipeline}

@@

Recently, Hariharan \etal. introduced Whitened Histograms of Orientations (WHO)
\cite{Hariharan12}, which uses statistics from natural images to whiten HOG
templates, making them more discriminative. Whitening is a common signal
processing operation for decorrelating a set of random variables
\cite{Martinsson05, Belouchrani00}. More formally, suppose that we have a
$k$-dimensional random variable $X \in \mathcal{R}^k$ with $Cov(X)=\Sigma$. By
whitening the signal,
\begin{equation}
\tilde{X}=\Sigma^{-\frac{1}{2}}(X - E[X]) \label{eq:whitening}
\end{equation}
, we
remove 2nd order correlation between all features. So, all elements in the
random variable $\tilde{X} = \left[\tilde{X}_1, \tilde{X}_2, \dots,
\tilde{X}_k\right]$, become more independent. In the context of HOG features, a
whitened HOG template will produce Gaussian noise when convolved with a random
natural image, but a delta-like response on an image matching the template. WHO
templates are more discriminative than standard HOG templates because they only
capture the gradients that differentiate the object from random images.

@@

We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), that are both more discriminative than WHO, and can be generated several orders of magnitude faster (in 70ms compared to several seconds for WHO). This speed up means we can generate templates on the fly, allowing our approach to evaluate arbitrary viewpoint hypotheses.

\subsubsection{Generating $\Sigma$ and $\mu$}
\scream{this section seems important to understand the following section but it's also just a re-writing of the Hariharan method}
In \cite{Hariharan12}, the authors compute the mean and covariance matrix by assuming Wide-Sense Stationarity (WSS) of HOG features
generated from natural images. That is, they assume the mean of a HOG cell is
independent of its place in the image, and the autocovariance of cells depends
only on their relative location.

In one dimension, let $x(u)$ be the feature at location $u$. Then WSS states
that $\mathbb{E}\left[x(u)\right] = \mu$ for all $u$, and
\begin{equation}
\textrm{cov}_x(u,v) = \textrm{cov}_x(0, v-u) = \Gamma(v-u),
\end{equation} where $\Gamma$ is the autocovariance. For simplicity, we describe the 1D case but this can be easily extended to 2D spatial autocovariance.

Therefore, assuming WSS allows the covariance matrix of templates \emph{of any size} to be synthesized from the autocovariance matrix using a simple lookup.

\subsubsection{Whitening Synthesized Templates}
Our first innovation is `Non-Zero' whitening. When synthesizing detection templates from rendered images, a common problem is how to handle the background. If the model is rendered over a natural image background, gradients in the background will be incorporated into the discriminative template, potentially causing false-positives if background elements exist in the query image.

Alternatively, if the background is left textureless \scream{did Hariharan do this?} (see Fig.~\ref{fig:rendering}), whitening the resulting HOG template will produce undesirable background artifacts. These are created when centering the template (by subtracting the mean $\mu$), where strong negative weights are introduced in the textureless region (as seen in Fig.~\ref{fig:whocomparison}). This could result in positive matches being suppressed due to spurious background gradients.

\scream{might need to match notation with subsequent sections}
NZ-WHO removes these artifacts so that the background has no effect on the template response. Given HOG features $x$, we create a new vector $x'$ which contains only the non-zero elements of $x$. We select the same elements of $\mu$ to produce $\mu'$, while the autocovariance lookup is only performed for the cells with non-zero gradient, producing $\Sigma'$ corresponding to the features in $x'$. After solving the resulting system (which is now smaller than in the WHO approach) we find
\begin{equation}
w'=\Sigma'^{-1}(x' - \mu') \label{eq:nz-who},
\end{equation}which we reassemble into the final NZ-WHO template $w$ by replacing the non-zero elements of $x$ with the corresponding $w'$ element.

Notice that $\Sigma$ is not square-rooted in Eq.~\ref{eq:nz-who}, whereas it is in the definition of whitening (Eq.~\ref{eq:whitening}). This is because the whitened template will only be used in the context of detection, which involves the inner product with the whitened HOG features of the query image $y$:
\begin{equation}
(y-\mu)^{T}\Sigma^{-\frac{1}{2}}\Sigma^{-\frac{1}{2}}(x-\mu)\\
= (y-\mu)^{T} w.
\end{equation}
We can therefore capture the whitening of the query image features in $w$.

% We use the GPU to parallelize the lookup of the correct autocovariance cells during covariance matrix syntheses, which is \scream{x} times faster than using the CPU (Figure \ref{fig:covariancetime}).

%\scream{chris:We propose Non-Zero Whitened Histograms of Orientations (NZ-WHO), an improvement over WHO which is both more discriminative than WHO and can be computed many orders of magnitude faster than WHO (70ms as opposed to to several seconds for WHO). This speedup allows us to generate NZ-WHO templates on the fly, allowing us to quickly evaluate arbitrary viewpoint hypotheses.}

\subsubsection{Fast Whitening}

@@

Table \ref{tab:who_initializations} provides a detailed comparison of WHO and NZ-WHO templates, illustrating that our method is faster and more discriminative than Whitened Histograms of Orientations. \scream{Does the table show this?}

% We compare the performance of WHO and NZ-WHO on 3D Object dataset \cite{Savarese07} and report performance on Table \ref{tab:who_initializations}.

@@

% \subsection{Whitening and Textureless Background}
\begin{comment}
\subsection{Constructing the Covariance Matrix}
\label{sec:feature_statistics}
% Whats used in our work and using Autocovariance matrix
We first collected the first and second order statistics (mean and
autocovariance) of HOG features from arbitrary natural images. Following
\cite{Hariharan12}, we assume Wide-Sense Stationarity (WSS) of HOG features
generated from natural images. That is, we assume the mean of a HOG cell is
independent of its place in the image, and the autocovariance of patches depends
only on their relative location.
